Motif roles provides unique information about species risk of extinction\\ \medskip Supplementary Information

### 2022-03-02 manuscript/response_PRSB_round2.tex

\them
Line 197 – 199: This sounds more statistically rigorous here than it does in the appendix where these are just visually compared.

\us Probably move this line to appendix

\them
Line 201: I have no idea what local and meso-scale refer to. Likewise global.

\us These terms are explained at their first use in the introduction or methods. To avoid confusion, we also add a glossary in \emph{Appendix S11} so that readers can easily look up these terms.

@@

\them
Line 200-206: The statistical analyses of the various components of this study seem very circular to me. Various local metrics depend on global metrics but treated differently in each step. There must be a more appropriate statistical technique for considering all of these moving parts in a comprehensive manner – even using model selection to compare.

\us See response to XX ... probably need to repeat that our focus is not to throw away degree

\them
Line 214: Why mean, and why not just that network persistence?

\us We use the term ``network mean persistence'' because this is the mean persistence for all species within the network. ``Network persistence'' could easily be misunderstood as referring to R50 or some other network-level measure of stability. To avoid confusion, we have added the term to the glossary in \emph{Appendix S11}

@@

\them
Line 259-262: This result seems odd to me. Is this just because all of the potential apparent competition interactions are filled first, and really is more of an artefact of the constraints about the types of motifs that are possible within these networks? And perhaps the connectances being explored are unrealistic given these constraints?

\us Empirical comparison should take care of this. ... citations to connectance range papers.

\them
Line 289: Odd use of direct competition when most other papers use exploitative competition

\us We are not aware of any other motif-focused papers calling this motif ``exploitative competition'' and follow the nomenclature from the introduction of motifs to food webs~\citep{Stouffer2007}.
Note that ``direct competition'' refers simply to two species which consume the same prey and is agnostic as to whether this is exploitative or interference competition.

@@

\them
Furthermore, the methods and statistical analysis were consistent with the proposed goals and very well written. I would like to highlight the correct use of motifs that authors made in the ms, which allows them to deduce new ecological information in addition to common network indices. Perhaps authors might include a figure with the used motifs in the appendix to facilitate the understanding about their structure and ecological meaning.

\us Same figure requested by Editor, good idea to add

\them
Regarding the results and discussion section, I found them very well written and fluent. However, because of the great number of results, I would suggest keeping consistency between the subtitles in the section methods and results referring to the same goal.

\us We have revised the subheadings to ensure consistency and thank the Reviewer for the reminder.

@@

\them
Line 240: Please keep consistency between the subtitles in the section methods and results referring to the same goal.

\us We have revised the subtitles to be consistent. Thank you for the suggestion.

\them
Line 270-272: Interesting. This could be related to the connectance of the species. Consumers with higher number of prey might have lower persistence at higher levels of disturbances because the disturbance has more pathways to propagate and reach the consumer.

\us We are not sure what the distinction between a species' degree and its connectance might be - we are used to connectance referring to network-level link density. 
However, we agree that this is an interesting result and had similar thoughts to the Reviewer that the interaction between degree and disturbance might indicate some kind of trade-off between the benefits of having many prey and the drawbacks of being exposed to disturbance through many pathways.
We had minimized this theme in the Discussion in order to meet word count limits, but have added a brief example in the final paragraph.

% Original line: In-degree shows an even stronger interaction with disturbance; having more prey species is associated with higher persistence without  disturbance, but lower persistence at high levels of disturbance.

### 2022-03-02 manuscript/supplement_motifs.tex
Updates from Overleaf

        To test whether networks with similar motif profiles tend to have similar mean persistences among consumers, we fit a PERMANOVA relating Bray-Curtis dissimilarity in motif profiles to the mean persistence of all consumers in a network (averaged across all levels of disturbance to basal resources).
        As with the PERMANOVA relating motif participation to network properties, we fit the PERMANOVA using the R~\citep{R} function `vegan' from the package \emph{adonis} and calculated significance based on 9999 permutations.

@@

    % Updated with new random effect, converted to data-scale
    \begin{table}[h!]
        \centering
        \caption{Coefficients ($\beta$) and $p$-values ($p$) for models relating persistence to the proportion of a species' role made up by the focal motif, the level of disturbance to basal resources (corresponding to a probability of extinction ranging between 0.1 and 0.5), and their interaction, as well as random effects to account for differences in persistence across levels of network size, connectance, and network ID  (Equation~\ref{propreq}). Predictors were centered and scaled before model fitting; coefficients refer to a unit increase in each parameter. We also provide the marginal $R^2_M$ (fixed effects only) and conditional $R^2_C$ (fixed and random effects) for each model.}
        \label{tab:proportion}                \footnotesize
        \begin{tabular}{c|c c | c c | c c | c c | c c |}
        & \multicolumn{2}{c|}{Intercept} & \multicolumn{2}{c|}{Proportion} & \multicolumn{2}{c|}{Disturbance} & \multicolumn{2}{c|}{Interaction} & \multicolumn{2}{c|}{$R^2$} \\
        Motif & $\beta$ & $p$ & $\beta$ & $p$ & $\beta$ & $p$ & $\beta$ & $p$ & $R^2_M$ & $R^2_C$ \\
        \hline
        3-sp Chain & 0.578 & \textless0.001 & -0.00360 & \textless0.001 & -0.0154 & \textless0.001 & 8.39$\times10^{-5}$& \textless0.001 & 0.866 & 0.914 \\
        App. Com. & 0.579 & \textless0.001 & -0.00239 & \textless0.001 & -0.0154 & \textless0.001 & -2.70$\times10^{-5}$ & \textless0.001 & 0.822 & 0.882 \\
        Dir. Com. & 0.577 & \textless0.001 & 7.74$\times10^{-4}$ & \textless0.001 & -0.0154 & \textless0.001 & -3.76$\times10^{-5}$ & \textless0.001 & 0.833 & 0.866 \\
        Omnivory & 0.577 & \textless0.001 & -0.00146 & \textless0.001 & -0.0154 & \textless0.001 & 4.22$\times10^{-5}$ & \textless0.001 & 0.842 & 0.877 \\
        \end{tabular}
    \end{table}        
\clearpage

@@

    % Updated with network-level random effect
    \begin{figure}[h!]
        \centering
        \includegraphics[height=.4\textheight]{figures/participation_vs_SC.eps}
        \caption{The proportion of each motif in a consumer's participation vector varied with network size, connectance, and their interaction. These relationships were steepest for apparent competition and omnivory.}
        \label{fig:roles_vs_SC}
    \end{figure}
\clearpage

@@

    % Updated with network-level random effect
    \begin{table}[h!]
        \caption{Coeffcients ($\beta$) for linear mixed-effect models of consumers' probabilities of persistence against simple measures of species' role (STL or in-degree), level of disturbance to basal resources, and their interaction, as well as a random effect of network ID. Predictors were centered and scaled before model fitting. All $p$-values were \textless0.001. The level of disturbance is reflected by basal resources having a baseline probability of extinction ranging between $0.1 - 0.5$.}
        \label{tab:per_vs_TLdeg}
        \centering
        \begin{tabular}{l|c  c  c |}
            Simple role measure & Role & Disturbance & Interaction \\
            \hline
            STL & -0.0210 & -0.139 & 0.00508 \\
            in-degree & -0.00188 & -0.139 & -0.00908 \\
        \end{tabular}
    \end{table}

    \begin{figure}[h!]
     \centering
     \includegraphics[width=.9\textwidth]{figures/heatmap_STL_allCS_BP0.pdf}
     \caption{All species across all levels experience a risk of going extinct due to causes not related to the web of $\pi = 0.1$. The sub figures show final persistence for consumers with different shortest trophic length, for increasing connectance (y-axis) and network size (x-axis). Persistence decreases with darker color in the heat map.}
     \label{fig:heatmap_stl_BP0}
    \end{figure}

    \begin{figure}[h!]
     \centering
     \includegraphics[width=.9\textwidth]{figures/heatmap_STL_allCS_BP1.pdf}
     \caption{Basal resources experience a disturbance increasing their probability to go extinct to $\pi = 0.5$. All consumer species go extinct due to causes not related to the web itself with probability $\pi = 0.1$. The sub figures show final persistence for consumers with different shortest trophic length, for increasing connectance (y-axis) and network size (x-axis). Persistence decreases with darker color in the heat map.}
     \label{fig:heatmap_stl_BP1}
    \end{figure}

@@

    \begin{figure}[h!]
        \centering
            \includegraphics[width=\textwidth]{figures/prop_dens_bp_vs_S_allC.pdf}
            \caption{Here we show the density (y-axis) of slopes (x-axis) of persistence against proportion of different motifs for all simulated webs of all connectances - a visualization of how an increased proportion of each motif (different colored lines) affects persistence of consumer species. Columns show the result for various disturbances on the basal level, from $\pi = 0.1$ (left) to $\pi = 0.5$ (right). Rows show various network sizes. The dotted, vertical lines indicate zero on the x-axis. A negative slope value reflects a negative relationship between increased participation in a motif and persistence, while a positive slope value reflects a positive relationship - an increased proportion of a specific motif increase persistence. The fraction of replicates with a slope greater than zero are stated in numbers in each sub-plot, the color corresponding to each motif (legend). }
            \label{fig:density_prop_S}
        \end{figure}

@@

    % Updated to use mean instead of all species
    \begin{table}[h!]
        \caption{Coefficients ($\beta$) for models relating persistence to the proportion of a network profile, the level of disturbance to basal resources (corresponding to a probability of extinction ranging between 0.1 and 0.5), and their interaction. Predictors were centered and scaled before model fitting. All effects were significant ($p$\textless0.05). We also provide the $R^2$ (multiple $R^2$). }
        \label{motif_profile_tab}
        \centering
        \begin{tabular}{c|c c c c c | c}
            Motif & Intercept & Proportion & Disturbance & Interaction & $R^2$\\
            \hline
            Omnivory & 0.574 & -0.0140 & -0.139 & -5.59$\times10^{-4}$ & 0.968 \\
            Apparent competition & 0.574 & 0.0112 & -0.139 & 5.90$\times10^{-4}$ & 0.964 \\
            Direct competition & 0.574 & 3.38$\times10^{-3}$ & -0.139 & 1.81$\times10^{-4}$ & 0.958 \\
            Three-species chain & 0.574 & 5.80$\times10^{-3}$ & -0.138 & -3.07$\times10^{-5}$ & 0.960 \\ 
        \end{tabular}
    \end{table}
\clearpage

@@

   \begin{figure}[h!]
       \centering
       \includegraphics[width=.75\textwidth]{figures/proportion_profile_dispersion.eps}
       \caption{Dispersion of normalized motif profiles (i.e., proportions of each motif) within a network about the centroid for that combination of network size and connectance. Each circle represents a single network; circle colour indicates network size. Scatter has been introduced about each level of connectance to increase visual clarity; the levels of connectance we consider are indicated by vertical black lines.}
       \label{dispersion_normmotifs}
    \end{figure}

@@

    % updated to use network means rather than all species
    \begin{figure}[h!]
        \centering
        \includegraphics[width=\textwidth]{figures/persistence_vs_SC_lm.eps}
        \caption{Persistence decreased strongly with increasing probability of disturbance to basal resources, while network size, connectance, and their interaction had smaller effects. \textbf{A)} Persistence decreased slightly with increasing network size at high levels of disturbance, increased slightly with increasing network size at low levels of disturbance, and did not vary noticeably with network size at moderate levels of disturbance. Networks with lower connectance (dotted line) had slightly higher persistence than networks with moderate (solid line) or high (dashed line) connectance.
        \textbf{B)} Persistence decreased with increasing connectance for all levels of disturbance and network size. Network size (line types) did not strongly affect these relationships.}
        \label{fig:lm_CS}
    \end{figure}

    \begin{figure}[h!]
      \includegraphics[width=\textwidth]{figures/heatmap_CS_BPcompare.eps}
       \caption{Heat maps of average consumer persistence in networks with varying connectance (y-axis) and network size (x-axis) at two levels of disturbance to basal resources. Persistence decreases with darker color in the heat map. When basal species are not disturbed, all species experience a baseline extinction risk of $\pi = 0.1$. When basal species are highly disturbed, their baseline extinction risk rises to $\pi = 0.5$ while the baseline risk of consumers is not changed.}
       \label{fig:heatmap_CS}
    \end{figure}

@@

 \begin{table}[h!]
 \label{glossary}
 \caption{Glossary of terms relating to motifs and Bayesian networks}
     \footnotesize{
 \begin{tabular}{l|l}
     Term & Definition \\
     \hline
     Motif & Set of $n$ interacting species. In this case, $n=3$ \\
     Network motif profile & Vector describing the frequency of each motif in the network. \\
     & Normalised by dividing counts of each motif by the total across all motifs.\\
     Motif participation role & Vector describing the frequency with which a focal species appears in each motif.\\
     & Normalised by dividing counts for each motif by the total across all motifs. \\
     Bayesian network & A directed acyclic graph, used to predict species' likelihood of persistence. \\
     Network persistence & The mean likelihood of consumers in a network not going extinct.\\
     Species persistence & The individual likelihood of a species not going extinct.\\
     In-degree & Number of prey species to a consumer species.\\
     Trophic level (STL) & The shortest food chain between the focal species and any basal species.\\
     Disturbance ($\pi_{disturbed}$) & Probability of extinction of a basal resource when extra disturbance is added. \\
     Baseline extinction &  When no disturbance is applied, $\pi_{base}$=0.1.\\
 \end{tabular}}
 \end{table}

>>>>

        To test whether networks with similar motif profiles tend to have similar mean persistences among consumers, we fit a PERMANOVA relating Bray-Curtis dissimilarity in motif profiles to the mean persistence of all consumers in a network (averaged across all levels of disturbance to basal resources).
        As with the PERMANOVA relating motif participation to network properties, we fit the PERMANOVA using the R~\citep{R} function `adonis' from the package \emph{vegan}~\citep{vegan} and calculated significance based on 9999 permutations.

@@

    % Updated with new random effect, converted to data-scale
    \begin{table}[ht!]
        \centering
        \caption{Coefficients ($\beta$) and $p$-values ($p$) for models relating persistence to the proportion of a species' role made up by the focal motif, the level of disturbance to basal resources (corresponding to a probability of extinction ranging between 0.1 and 0.5), and their interaction, as well as random effects to account for differences in persistence across levels of network size, connectance, and network ID  (Equation~\ref{propreq}). Predictors were centered and scaled before model fitting; coefficients refer to a unit increase in each parameter. We also provide the marginal $R^2_M$ (fixed effects only) and conditional $R^2_C$ (fixed and random effects) for each model.}
        \label{tab:proportion}                \footnotesize
        \begin{tabular}{c|c c | c c | c c | c c | c c |}
        & \multicolumn{2}{c|}{Intercept} & \multicolumn{2}{c|}{Proportion} & \multicolumn{2}{c|}{Disturbance} & \multicolumn{2}{c|}{Interaction} & \multicolumn{2}{c|}{$R^2$} \\
        Motif & $\beta$ & $p$ & $\beta$ & $p$ & $\beta$ & $p$ & $\beta$ & $p$ & $R^2_M$ & $R^2_C$ \\
        \hline
        3-sp Chain & 0.578 & \textless0.001 & -0.00360 & \textless0.001 & -0.0154 & \textless0.001 & 8.39$\times10^{-5}$& \textless0.001 & 0.866 & 0.914 \\
        App. Com. & 0.579 & \textless0.001 & -0.00239 & \textless0.001 & -0.0154 & \textless0.001 & -2.70$\times10^{-5}$ & \textless0.001 & 0.822 & 0.882 \\
        Dir. Com. & 0.577 & \textless0.001 & 7.74$\times10^{-4}$ & \textless0.001 & -0.0154 & \textless0.001 & -3.76$\times10^{-5}$ & \textless0.001 & 0.833 & 0.866 \\
        Omnivory & 0.577 & \textless0.001 & -0.00146 & \textless0.001 & -0.0154 & \textless0.001 & 4.22$\times10^{-5}$ & \textless0.001 & 0.842 & 0.877 \\
        \end{tabular}
    \end{table}        
\clearpage

@@

    % Updated with network-level random effect
    \begin{figure}[ht!]
        \centering
        \includegraphics[height=.4\textheight]{figures/participation_vs_SC.eps}
        \caption{The proportion of each motif in a consumer's participation vector varied with network size, connectance, and their interaction. These relationships were steepest for apparent competition and omnivory.}
        \label{fig:roles_vs_SC}
    \end{figure}
\clearpage

@@

    % Updated with network-level random effect
    \begin{table}[hb!]
        \caption{Coeffcients ($\beta$) for linear mixed-effect models of consumers' probabilities of persistence against simple measures of species' role (STL or in-degree), level of disturbance to basal resources, and their interaction, as well as a random effect of network ID. Predictors were centered and scaled before model fitting. All $p$-values were \textless0.001. The level of disturbance is reflected by basal resources having a baseline probability of extinction ranging between $0.1 - 0.5$.}
        \label{tab:per_vs_TLdeg}
        \centering
        \begin{tabular}{l|c  c  c |}
            Simple role measure & Role & Disturbance & Interaction \\
            \hline
            STL & -0.0210 & -0.139 & 0.00508 \\
            in-degree & -0.00188 & -0.139 & -0.00908 \\
        \end{tabular}
    \end{table}

    \begin{figure}[hb!]
     \centering
     \includegraphics[width=.9\textwidth]{figures/heatmap_STL_allCS_BP0.pdf}
     \caption{All species across all levels experience a risk of going extinct due to causes not related to the web of $\pi = 0.1$. The sub figures show final persistence for consumers with different shortest trophic length, for increasing connectance (y-axis) and network size (x-axis). Persistence decreases with darker color in the heat map.}
     \label{fig:heatmap_stl_BP0}
    \end{figure}

    \begin{figure}[hb!]
     \centering
     \includegraphics[width=.9\textwidth]{figures/heatmap_STL_allCS_BP1.pdf}
     \caption{Basal resources experience a disturbance increasing their probability to go extinct to $\pi = 0.5$. All consumer species go extinct due to causes not related to the web itself with probability $\pi = 0.1$. The sub figures show final persistence for consumers with different shortest trophic length, for increasing connectance (y-axis) and network size (x-axis). Persistence decreases with darker color in the heat map.}
     \label{fig:heatmap_stl_BP1}
    \end{figure}

@@

    \begin{figure}[hb!]
        \centering
            \includegraphics[width=\textwidth]{figures/prop_dens_bp_vs_S_allC.pdf}
            \caption{Here we show the density (y-axis) of slopes (x-axis) of persistence against proportion of different motifs for all simulated webs of all connectances - a visualization of how an increased proportion of each motif (different colored lines) affects persistence of consumer species. Columns show the result for various disturbances on the basal level, from $\pi = 0.1$ (left) to $\pi = 0.5$ (right). Rows show various network sizes. The dotted, vertical lines indicate zero on the x-axis. A negative slope value reflects a negative relationship between increased participation in a motif and persistence, while a positive slope value reflects a positive relationship - an increased proportion of a specific motif increase persistence. The fraction of replicates with a slope greater than zero are stated in numbers in each sub-plot, the color corresponding to each motif (legend). }
            \label{fig:density_prop_S}
        \end{figure}

@@

    % Updated to use mean instead of all species
    \begin{table}[hb!]
        \caption{Coefficients ($\beta$) for models relating persistence to the proportion of a network profile, the level of disturbance to basal resources (corresponding to a probability of extinction ranging between 0.1 and 0.5), and their interaction. Predictors were centered and scaled before model fitting. All effects were significant ($p$\textless0.05). We also provide the $R^2$ (multiple $R^2$). }
        \label{motif_profile_tab}
        \centering
        \begin{tabular}{c|c c c c c | c}
            Motif & Intercept & Proportion & Disturbance & Interaction & $R^2$\\
            \hline
            Omnivory & 0.574 & -0.0140 & -0.139 & -5.59$\times10^{-4}$ & 0.968 \\
            Apparent competition & 0.574 & 0.0112 & -0.139 & 5.90$\times10^{-4}$ & 0.964 \\
            Direct competition & 0.574 & 3.38$\times10^{-3}$ & -0.139 & 1.81$\times10^{-4}$ & 0.958 \\
            Three-species chain & 0.574 & 5.80$\times10^{-3}$ & -0.138 & -3.07$\times10^{-5}$ & 0.960 \\ 
        \end{tabular}
    \end{table}
\clearpage

@@

   \begin{figure}[hb!]
       \centering
       \includegraphics[width=.75\textwidth]{figures/proportion_profile_dispersion.eps}
       \caption{Dispersion of normalized motif profiles (i.e., proportions of each motif) within a network about the centroid for that combination of network size and connectance. Each circle represents a single network; circle colour indicates network size. Scatter has been introduced about each level of connectance to increase visual clarity; the levels of connectance we consider are indicated by vertical black lines.}
       \label{dispersion_normmotifs}
    \end{figure}

@@

    % updated to use network means rather than all species
    \begin{figure}[hb!]
        \centering
        \includegraphics[width=\textwidth]{figures/persistence_vs_SC_lm.eps}
        \caption{Persistence decreased strongly with increasing probability of disturbance to basal resources, while network size, connectance, and their interaction had smaller effects. \textbf{A)} Persistence decreased slightly with increasing network size at high levels of disturbance, increased slightly with increasing network size at low levels of disturbance, and did not vary noticeably with network size at moderate levels of disturbance. Networks with lower connectance (dotted line) had slightly higher persistence than networks with moderate (solid line) or high (dashed line) connectance.
        \textbf{B)} Persistence decreased with increasing connectance for all levels of disturbance and network size. Network size (line types) did not strongly affect these relationships.}
        \label{fig:lm_CS}
    \end{figure}

    \begin{figure}[hb!]
      \includegraphics[width=\textwidth]{figures/heatmap_CS_BPcompare.eps}
       \caption{Heat maps of average consumer persistence in networks with varying connectance (y-axis) and network size (x-axis) at two levels of disturbance to basal resources. Persistence decreases with darker color in the heat map. When basal species are not disturbed, all species experience a baseline extinction risk of $\pi = 0.1$. When basal species are highly disturbed, their baseline extinction risk rises to $\pi = 0.5$ while the baseline risk of consumers is not changed.}
       \label{fig:heatmap_CS}
    \end{figure}

@@

 \begin{table}[hb!]
 \label{glossary}
 \caption{Glossary of terms relating to motifs and Bayesian networks}
     \footnotesize{
 \begin{tabular}{l|m{11cm}}
     Term & Definition \\
     \hline
     Global property & A property of the whole network (e.g., species richness) \\
     Local property & A property of a single species, generally referring only to direct interactions (e.g., degree) \\
     Meso-scale property &  A property of a single species which includes direct and indirect interactions; the species' immediate `neighbourhood' in the network (e.g., motif participation) \\
     Motif & Set of $n$ interacting species. In this case, $n=3$ \\
     Network motif profile & Vector describing the frequency of each motif in the network. \\
     & Normalised by dividing counts of each motif by the total across all motifs.\\
     Motif participation role & Vector describing the frequency with which a focal species appears in each motif.\\
     & Normalised by dividing counts for each motif by the total across all motifs. \\
     Bayesian network & A directed acyclic graph, used to predict species' likelihood of persistence. \\
     Network mean persistence & The mean likelihood of persistence for all consumers in a network.\\
     Species persistence & The individual likelihood of a species not going extinct.\\
     In-degree & Number of prey species to a consumer species.\\
     Trophic level (STL) & The shortest food chain between the focal species and any basal species.\\
     Disturbance ($\pi_{disturbed}$) & Probability of extinction of a basal resource when extra disturbance is added. \\
     Baseline extinction &  When no disturbance is applied, $\pi_{base}$=0.1.\\
 \end{tabular}}
 \end{table}
